---
title: "Project_1"
author: "Nikhileswar mada"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 1

*Implement the Bisection method. Design a test problem with a known root x0. Apply the
Bisection method to the test problem with a tolerance $\epsilon=10^{-6}$.Generate a table to report the iteration number n.The value of $n^{th}$ approximation is $x_n$,the function value $f(x_n)$,the error $e_n=|x_n-x_0|$, and the ratio $\frac{e_n}{e_{n-1}}$.*

Ans:
We are choosing $f(x)=x^3-3x+1$ as our equation and [a,b]=[1,2]
as f(a)<0 and f(b)>0
```{r}
fun<-function(x) x^3 - 3*x + 1
a<-1
b<-2
tol<-10^-6
n<-0
x_n<-(a+b)/2
e_n<-abs(x_n-1.87938524)
e_frac<-0
answer<-data.frame(n, x_n, fun(x_n), e_n, e_frac)
while(abs(fun(x_n)) > tol){
  if (fun(x_n)==0){
    break
  }
  else if(fun(a)*fun(x_n) < 0){
    b <- x_n
  } else {
    a <- x_n
  }
  n <- n + 1
  x_n_prev <- x_n
  x_n <- (a+b)/2
  e_n_prev <- e_n
  e_n <- abs(x_n - 1.87938524)
  e_frac <- e_n/e_n_prev
  answer <- rbind(answer, c(n, x_n, fun(x_n), e_n, e_frac))
}
print(answer)
```
Above shows the table for the required factors using bisection method after 20 iterations we have achieved the function value at x=1.532089 be less than tolerance of $10^{-6}$.

\newpage

# Question 2

Known root of $x^3-6=0$ between 1 and 2 is obtained from internet in order to calculate the error $|x_n-x_0|$ which is $x_0=1.8171205928321397$
Using the bisection first to determine the initial guess for Newton's method.
f(1)<0 and f(2)>0.So,our initial guess for [a,b]=[1,2]

```{r}
fun<-function(x) x^3 - 6
a<-1
b<-2
x_n<-(a+b)/2
tol_b<-10^-3
while(abs(fun(x_n))>tol_b){
  if(fun(x_n)==0){
    break
  }
  else if(fun(a)*fun(x_n) < 0){
    b <- x_n
  } else {
    a <- x_n
  }
  x_n <- (a+b)/2
}

x_newton<-x_n
fun_dev<-function(x) 3*x^2
tol_n<-10^-10
n<-0
x_newton_prev<-Inf
error<-abs(x_newton-1.8171205928321397)
practical_error_estimator<-Inf
answer<-data.frame(n,x_newton,error,practical_error_estimator)
while(abs(fun(x_newton))>tol_n){
  x_newton_prev<-x_newton
  x_newton<-x_newton-fun(x_newton)/fun_dev(x_newton)
  n<-n+1
  error<-abs(x_newton-1.8171205928321397)
  practical_error_estimator<-abs(x_newton-x_newton_prev)/abs(x_newton_prev)
  answer<-rbind(answer,c(n,x_newton,error,practical_error_estimator))
}
print(answer)
```
Above shows the table for the required factors using newton's method after 3 iterations we have achieved the function value at x=1.817121 be less than tolerance of $10^{-10}$.

\newpage

# Question 3

given $f(x)=0$ (i.e) $3x-sin(x)=0$ We know that fixed point problem

$$x=g(x) \implies f(x)=0$$
$$x=1/3(sin(x))$$
$$g(x)=1/3(sin(x))$$



then $$x_{n+1}=g(x_n)$$



We know that the $g^{'}(x)$ is continuous and clearly 0 is a fixed point of g(x) and $|g^{'}(r)<1|$,this implies that the point iteration is locally convergent.So,we can the point iteration method.
let's start x=0.5 radians.
```{r}
fun<-function(x)1/3*(sin(x))
x_n<-0.5
tol<-10^-6
n<-0
x_n_prev<-Inf
practical_error_estimator<-Inf
error<-abs(0.5-0)
answer<-data.frame(n,x_n,error,practical_error_estimator)
while(abs(fun(x_n)-x_n)>tol){
  x_n_prev<-x_n
  x_n<-fun(x_n)
  n<-n+1
  error<-abs(x_n-0)
  practical_error_estimator<-abs(x_n-x_n_prev)/abs(x_n_prev)
  answer<-rbind(answer,c(n,x_n,error,practical_error_estimator))
}

print(answer)
```

Above shows the table for the required factors using convergent fixed point iteration after 12 iterations we have achieved the function value at x=$8.97*10^{-7}$ be less than tolerance of $10^{-6}$.

\newpage

# Question 4

since we don't know the [a,b] intutively let's try and find the bounds using substitution
```{r}

fun<-function(x) (x-1)*(x-2)*(x-3)*(x-4)*(x-5)*(x-6)*(x-7)-(10^-7)*x^7
a<-6.5
b<-7.5
print(fun(a))
print(fun(b))
```

As we can see that $f(a).f(b)<0$.So,our guess of [a,b]=[6.5,7.5] can be used

```{r}
tol<-10^-8
n<-0
x_n<-(a+b)/2
e_n<-abs(x_n-7)
e_frac<-0
answer<-data.frame(n, x_n, fun(x_n), e_n, e_frac)
while(abs(fun(x_n)) > tol){
  if (fun(x_n)==0){
    break
  }
  else if(fun(a)*fun(x_n) < 0){
    b <- x_n
  } else {
    a <- x_n
  }
  n <- n + 1
  x_n_prev <- x_n
  x_n <- (a+b)/2
  e_n_prev <- e_n
  e_n <- abs(x_n - 7)
  e_frac <- e_n/e_n_prev
  answer <- rbind(answer, c(n, x_n, fun(x_n), e_n, e_frac))
}
print(answer)
```

After 35 iterations the value of function is less than tolerance of $10^{-8}$ at x=7.000114.

