---
title: "Project2"
author: "Nikhileswar mada"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning=FALSE)
```

# Question 1

implementing the pre conditioner function using forward and backward substitution

```{r}
preconditioner<-function(A,r){
  D<-diag(diag(A))
  L<-lower.tri(A)
  U<-upper.tri(A)
  c<-forward(D+L,r)
  z<-backward(solve(D) %*% (D+U),c)
  return(z)
}

forward<-function(L,b){
  n<-nrow(L)
  x<-rep(0,n)
  for (i in 1:n){
    
    x[i]<-(b[i]-L[i,1:(i-1)] %*% x[1:(i-1)])/L[i,i]
  }
  return(x)
}

backward<-function(U,c){
  n<-nrow(U)
  x<-rep(0,n)
  for (i in n:1){
    if (i==n){
      x[i]<-(c[i]/U[i,i])
    }
    else{
    x[i]<-(c[i]-U[i,(i+1):n] %*%x[(i+1):n])/U[i,i]
    }
  }
  return(x)
}

```

sample to check whether the functions are working as intended

We need a to be symmetric and positive definite
```{r}
n <- 10
A <- matrix(rnorm(n^2), n, n)
A <- A %*% t(A) 
r <- rnorm(n)
#LU<-lu(A)
z<-solve(A,r)
x<-preconditioner(A,r)
#elu<-expand(LU)
#L<-elu$L
#P<-elu$P
#U<-elu$U
#y<-solve(L,solve(P)%*% r)
#x<-solve(U,y)
x
z
```


We found the solutions are equal.

\newpage

# Question 2


Reading A using read table.

```{r}
A<-read.table('A.dat',sep=',')
A<-as.matrix(A)
```

```{r}
n<-100
xt<-rep(1,nrow(A))
b<-A%*%xt
tol<-1e-10
residual_without<-rep(0,n)
residual_with<-rep(0,n)
x0<-rep(3,nrow(A))
r<-b-A%*%x0
x<-x0
error_with_preconditioner<-rep(0,n)
error_without_preconditioner<-rep(0,n)
for (i in 1:n){
  z<-r
  #z_with<-preconditioner(A,r)
  alpha<-(t(z)%*%r)/(t(z)%*%A%*%z)
  residual_without[i]<-sqrt(sum((b-A%*%x)^2))/sqrt(sum(b^2))
  if(residual_without[i]<tol){
    break
  }
  r<-r- as.numeric(alpha)*A%*%z
  x<-x+as.numeric(alpha)*z
}


r<-b-A%*%x0
x<-x0
for (j in 1:n){
  z<-preconditioner(A,r)
  alpha<-(t(z)%*%r)/(t(z)%*%A%*%z)
  residual_with[j]<-sqrt(sum((b-A%*%x)^2))/sqrt(sum(b^2))
  if(residual_with[j]<tol){
    break
  }
  r<-r- as.numeric(alpha)*A%*%z
  x<-x+as.numeric(alpha)*z  
}
# Saving the residuals and the iterations into RData file.
save(residual_with,file='question2.RData')
save(residual_without,file='question2.RData')
plot(exp(residual_without[1:100]),type='l',col='red',ylab='Residual',xlab='Iterations')
lines(exp(residual_with),col='green')
legend(x='topright',legend=c('without preconditioner','with preconditioner'),col=c('green','red'),lwd=2)
```

Saving the Residuals of both with and without the use of preconditioner into question.RData file.

\newpage

# Question 3

We repeat the same thing here as the question 2


```{r}
n<-1000
xt<-rep(1,nrow(A))
b<-A%*%xt
tol<-1e-10
residual_without<-rep(0,n)
x0<-rep(3,nrow(A))
r<-b-A%*%x0
x<-x0
z=r
p=z
for (i in 1:n){
  
  #z_with<-preconditioner(A,r)
  alpha<-(t(p)%*%r)/(t(p)%*%A%*%z)
  residual_without[i]<-sqrt(sum((b-A%*%x)^2))/sqrt(sum(b^2))
  if(residual_without[i]<tol){
    break
  }
  r<-r-as.numeric(alpha)*A%*%p
  x<-x+as.numeric(alpha)*p
  z<-r
  beta<-(t(p)%*%A%*%z)/(t(p)%*%A%*%p)
  p<-z-as.numeric(beta)*p
}
plot(residual_without,type='l')

```

The loop broke after 87 iterations without preconditioning.

```{r}
n<-100
r<-b-A%*%x0
x<-x0
z=preconditioner(A,r)
p=z
residual_with<-rep(0,n)
for (i in 1:n){
  #z_with<-preconditioner(A,r)
  alpha<-(t(p)%*%r)/(t(p)%*%A%*%z)
  residual_with[i]<-sqrt(sum((b-A%*%x)^2))/sqrt(sum(b^2))
  if(residual_with[i]<tol){
    break
  }
  r<-r-as.numeric(alpha)*A%*%p
  x<-x+as.numeric(alpha)*p
  z<-preconditioner(A,r)
  beta<-(t(p)%*%A%*%z)/(t(p)%*%A%*%p)
  p<-z-as.numeric(beta)*p
}
save(residual_with,file='question3.RData')
save(residual_without,file='question3.RData')
plot(exp(residual_with),type='l',col='green',ylab='residual',xlab='Iterations')
lines(exp(residual_without),col='red')
legend(x='topright',legend=c('without preconditioner','with preconditioner'),col=c('green','red'),lwd=2)
```


We can see in both the cases that using preconditioner quickly dies down the loss 







